A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-05-22 paper/paper1/ms.tex
More CVM text edits from draft markup

We can constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range may be typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \kms---assuming a circular orbit---slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  Notably, our fit does not jointly model the Doppler tomographic signals reported in \citep{2017AJ....153..211Z}, which could further constrain the impact parameter.  Additionally, our input $\rstar$ was slightly lower than the best-fit value derived from \emph{Gaia DR3} and SED fitting.
\authorcomment1{Quang Tran: please verify and revise.}

\subsection{HPF analysis II: \ion{He}{1} 10833 \AA} \label{secHeAnalysis}
The preparation for analyzing the Helium feature followed a similar but slightly different procedure than that used for the RV, namely, we used \texttt{Goldilocks}\footnote{\url{https://github.com/grzeimann/Goldilocks_Documentation}} for 2D echellogram reduction.  This tool outputs 1D extracted spectra for the target, and two reference fibers: blank sky and a laser frequency comb \citep[LFC,][]{2019Optic...6..233M}.  The observations were usually acquired with the LFC turned off, so this unilluminated spectrum was discarded.

@@

Excess is confidently detected from phases $-$0.3 to $+$0.03.  The full-width at half maximum (FWHM) of the feature is about 1.8-2.6~\AA, equivalent to a line-of-sight velocity distribution in the range of 50-75~km/s.  Individual fits to the He feature show a typical line center uncertainty $\pm0.1$~\AA~ or better.  The out-of-transit phases exhibit a line center position of 10833.4~\AA, consistent with the stellar \ion{He}{1} rest frame zero velocity.  The excess absorption resides slightly blueward of the stellar zero restframe velocity reference.  As mentioned previously, the telluric masking near 10836 \AA~ censors our view of the existence (or not) of any redshifted lobe in this region.  A more careful treatment of the telluric absorption could rescue some of the data in this region.  The excess detections exhibit a bulk blueshift of up to 30~km/s.

@@

\subsection{Size and geometry of outflow}
Helium absorption occurs predominantly before the planet is in-transit, indicating a \emph{leading} tail escaping the planet.  The excess occurs as early as 30\% of the orbital path.  That large fraction of an orbit translates to about 0.1 AU, 130 planetary radii, and well outside of the Roche Lobe, with its Hill radius of merely 2.6 planetary radii.  We are detecting Roche lobe overflow, escaping material that is entirely unbound from the planet.  The bulk of the detectable unbound gas may be understood as following its own Keplerian trajectory at a slightly shorter period orbit, pulling away from the planet parallel to the orbital path.  Such a scenario may arise from preferential emission on the planet dayside.

The large size of the gas stream overfills the transit chord, which means both the advancing blue side and receding red side of the rotating star are nearly continuously obscured in this spin-orbit-aligned system \citep{2017AJ....153..211Z}. The planet transits near the Stellar equator \citep{2017AJ....153..211Z}, meaning the transit chord covers about 10$\%$ of the projected stellar surface area.  Hypothetically a nearly 100$\%$ opaque optically thick stream of gas covering only the transit chord could reproduce the observed 10$\%$ absorption excess.  Alternatively---and more likely---an optically thin gas stream has some additional vertical extent perpendicular to the plane of orbit, which would act to overfill the entire projected stellar disk.  Assuming the gas has a spatially uniform optical depth of about 0.1 would reproduce the 10\% absorption excess.

\subsection{1D Parker wind models and mass loss rates}\label{pwinds}
The dramatic leading tail of atmospheric escape suggests that more complicated 3D hydrodynamic simulations coupled with stellar wind \citep{2022ApJ...926..226M} may be needed to depict the geometry faithfully.  Here we explore one-dimensional (1D) models, which offer ease of interpretation and rapid calculation.  We employ the open-source \texttt{p-winds} code \citep{2022A&A...659A..62D}, which implements a transonic 1D Parker wind model with radiative transfer of the Helium 10830 triplet \citep{2018ApJ...855L..11O,2020A&A...636A..13L}.

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally, we would have a measurement of the XUV spectrum of \hatp, but the limited facilities and distance of the source preclude these challenging observations.  Instead, we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with available synthetic X-ray coronal spectra \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with weakening of the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for HAT-P-67 \citep{2022ApJ...930....7A}.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} XUV spectrum to HAT-P-67.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-6}, 10^{-4}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with these high and low radiation hardnesses.

@@

At least a few shortcomings limit the applicability of this 1D Parker wind model.  First, the model-dependent line profile (\emph{i.e. width and depth}) is degenerate with XUV flux, $\dot{M}$, and $T_0$, and so a range of these parameters can be fine-tuned to obtain a large range of mass loss rates consistent with the data and our limited understanding of the XUV flux.  These known degeneracies have been pointed out previously \citep{2022AJ....164..234V,2019ApJ...881..133O}, but the problem for \hatpb~ appears somewhat more acute since XUV data are scarce for F-type stars near the Kraft break.  Second, the 1D model breaks down when attempting to explain the inherently 3D leading tail geometry.

@@

\subsection{HAT-P-67 b analogues}

>>>>

We can constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \kms---assuming a circular orbit---slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.
\authorcomment1{Quang Tran: please verify and revise.}

\subsection{HPF analysis II: \ion{He}{1} 10833 \AA} \label{secHeAnalysis}
The preparation for analyzing the Helium feature followed a similar but slightly different procedure than that used for the RV, namely, we used \texttt{Goldilocks}\footnote{\url{https://github.com/grzeimann/Goldilocks_Documentation}} for 2D echellogram reduction.  This tool outputs 1D extracted spectra for the target, and two reference fibers: blank sky and a laser frequency comb \citep[LFC,][]{2019Optic...6..233M}.  The observations were acquired with the LFC turned off, so this unilluminated spectrum was discarded.

@@

Excess is confidently detected from phases $-$0.3 to $+$0.03.  The full-width at half maximum (FWHM) of the feature is about 1.8-2.6~\AA, equivalent to a line-of-sight velocity distribution in the range of 50-75~km/s.  Individual fits to the He feature show a typical line center uncertainty $\pm0.1$~\AA~ or better.  The out-of-transit phases exhibit a line center position of 10833.4~\AA, consistent with the stellar \ion{He}{1} rest frame zero velocity.  The excess absorption resides slightly blueward of the stellar zero restframe velocity reference.  As mentioned previously, the telluric masking near 10836 \AA~ censors our view of the existence (or not) of any redshifted lobe in this region.  The excess detections exhibit a bulk blueshift of up to 30~km/s.

@@

\subsection{Size and geometry of outflow}
Helium absorption occurs predominantly before the planet is in-transit, indicating a \emph{leading} tail escaping the planet.  The leading tail is detected 0.1 AU away from the planet, or 130 planetary radii, well outside of the Roche Lobe at $<2.6$ planetary radii.  We are detecting Roche lobe overflow, escaping material that is entirely unbound from the planet.  The bulk of the detectable unbound gas may be understood as following its own Keplerian trajectory at a slightly shorter period orbit, pulling away from the planet parallel to the orbital path.  Such a scenario may arise from preferential emission on the planet dayside.

The large size of the gas stream overfills the transit chord, which means both the advancing blue side and receding red side of the rotating star are nearly continuously obscured in this spin-orbit-aligned system \citep{2017AJ....153..211Z}. The planet transits near the stellar equator \citep{2017AJ....153..211Z}, meaning the transit chord covers about 10$\%$ of the projected stellar surface area.  Hypothetically a nearly 100$\%$ opaque optically thick stream of gas covering only the transit chord could reproduce the observed 10$\%$ absorption excess.  Alternatively---and more likely---an optically thin gas stream has some additional vertical extent perpendicular to the plane of orbit, which would act to overfill the entire projected stellar disk.  Assuming the gas has a spatially uniform optical depth of about 0.1 would reproduce the 10\% absorption excess.

\subsection{1D Parker wind models and mass loss rates}\label{pwinds}
First we explore one-dimensional (1D) models, which offer ease of interpretation and rapid calculation.  We employ the open-source \texttt{p-winds} code \citep{2022A&A...659A..62D}, which implements a transonic 1D Parker wind model with radiative transfer of the Helium 10830 triplet \citep{2018ApJ...855L..11O,2020A&A...636A..13L}.

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally, we would have a measurement of the XUV spectrum of HAT-P-67, but the limited facilities and distance of the source preclude these challenging observations.  Instead, we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with available synthetic X-ray coronal spectra \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with weakening of the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for HAT-P-67 \citep{2022ApJ...930....7A}.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} XUV spectrum to HAT-P-67.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-6}, 10^{-4}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with these high and low radiation hardnesses.

@@

At least a few shortcomings limit the applicability of this 1D Parker wind model.  First, the model-dependent line profile (\emph{i.e.} width and depth) is degenerate with XUV flux, $\dot{M}$, and $T_0$, and so a range of these parameters can be fine-tuned to obtain a large range of mass loss rates consistent with the data and our limited understanding of the XUV flux.  These known degeneracies have been pointed out previously \citep{2022AJ....164..234V,2019ApJ...881..133O}, but the problem for \hatpb~ appears somewhat more acute since XUV data are scarce for F-type stars near the Kraft break.  Second, the 1D model breaks down when attempting to explain the inherently 3D leading tail geometry.

@@

\subsection{Other planets likely to be evaporating}
